Normalization by Evaluation for Call-By-Push-Value and Polarized Lambda Calculus

### 2019-05-07 ppdp19/ppdp19.tex
[ ppdp19 ] draft more for hyperref no longer needed
[ ppdp19 ] proof reading until half of 2.3

\hypersetup{draft}

@@

%
% The "author" command and its associated commands are used to define the authors and their affiliations.
% Of note is the shared affiliation of the first two authors, and the "authornote" and "authornotemark" commands
% used to denote shared contribution to the research.
\author{Andreas Abel}
%\authornote{Both authors contributed equally to this research.}
\email{andreas.abel@gu.se}
\orcid{0000-0003-0420-4492}
%\authornotemark[1]
\affiliation{%
  \institution{Department of Computer Science and Engineering, Chalmers and Gothenburg University}
  \streetaddress{Rännvägen 6b}
  \city{Göteborg}
  \state{Sweden}
  \postcode{41296}
}

\author{Christian Sattler}
\affiliation{%
  \institution{Nottingham University}
  \streetaddress{}
  \city{Nottingham}
  \country{UK}
}
\email{christian.sattler@gmail.com}

@@

The idea behind \emph{normalization by evaluation} (NbE) is to utilize a
standard interpreter, usually evaluating closed terms, to compute the
normal form of an open term.
The normal form is obtained by a type-directed \emph{reification}
procedure after evaluating the open term to a semantic value, mapping
(\emph{reflecting})
the free variables to corresponding \emph{unknowns} in the semantics.
The literal use of a standard interpreter
can be achieved for the pure simply-typed lambda-calculus
\cite{bergerSchwichtenberg:lics91,filinski:semaccounttdpe}
by modelling uninterpreted base types as sets of neutral (aka atomic)
terms, or more precisely, as presheaves or sets of neutral term
families, in order to facilitate fresh bound variable generation during
reification of functions to lambdas.
Thanks to $\eta$-equality at function types, free variables of
function type can be reflected into the semantics as functions
applying the variable to their reified argument,
forming a neutral term.  This mechanism provides us with unknowns of
function type which can be faithfully reified to normal forms.
% Since neutrals are closed under application to normal arguments, they
% can act as unknown functions, storing the reified semantic argument
% they are invoked with (thus, these unknowns are aka accumulators
% \cite{gregoireleroy:icfp02}).

For the extension to sum types (logically, disjunctions, and categorically,
weak coproducts), this reflection trick does not work anymore.  A
semantic value of binary sum type is either a left or a right
injection, but the decision between left or right cannot be taken at
reflection time, since a variable of sum type does not provide us with
such information.  A literal standard interpreter for closed terms can
thus no longer be used for NbE;  instead, we can utilize a monadic
interpreter.  When the interpreter attempts a case distinction on an
unknown of sum type, it asks an oracle whether the unknown is a left
or a right injection.  The oracle returns one of these alternatives,
wrapping a new unknown in the respective injection.  The communication with the
oracle can be modeled in a \emph{monad} $\C$,
% with the service $\tNe\,(A + B) \to \C\,(\den A + \den B)$
which records the questions asked and the continuation of the
interpreter for each of the possible answers.
A monadic semantic value is thus a \emph{case tree} where the leaves
are occupied by non-monadic values \cite{altenkirchUustalu:flops04}.
In this article we only consider weak sum types, producing non-unique
normal forms, where it does not
matter in which order the questions are asked (commuting case splits),
and whether the same
question is asked several times (redundant case splits).
The model would need refinement for strong, extensional sums
\cite{altenkirchDybjerHofmannScott:lics01,altenkirchUustalu:flops04,balatDiCosmoFiore:popl04,barral:PhD,scherer:popl17}.

@@

In this work, we formulate NbE for CBPV
(Section~\ref{sec:cbpv}), with the aim to investigate
later whether CBN and CBV NbE can be recovered from CBPV NbE via the
standard translations of the CBN and CBV calculi into CBPV.

In contrast to the normal forms of CBN NbE, which is the algorithmic
counterpart of the completeness proof for intuitionistic propositional
logic (IPL) using Beth models, CBPV NbE gives us more restrained
normal forms, where the production of a value via injections cannot be
interrupted by more questions to the oracle.  In the research field of
focalization \cite{andreoli:focusing,liangMiller:csl07} we speak of
\emph{chaining non-invertible introductions}.
Invertible introductions are already chained in NbE thanks to
extensionality ($\eta$) for function, and more generally, negative
types.  Non-invertible eliminations are also happening in a chain when
building neutrals.  What is missing from the picture is the chaining
of invertible eliminations, i.e., case distinctions and, more
generally, pattern matching.  The picture is completed by extending
NbE to \emph{polarized lambda calculus}
\cite{zeilberger:PhD,brockNannestadSchuermann:lpar10,espiritoSanto:entcs17}
in Section~\ref{sec:fipl}.

@@

By a \emph{presheaf} we will always mean a presheaf over the category $\Cxt\op$,
i.e., a functor from $\Cxt$ to $\Set$.\footnote{With respect to the terms
defined later, $\Cxt\op$ is to be seen as a category of contexts and those
substitutions given by weakening.}  For example, given any type $A$, we can
consider $A{\in}\_$ as such a presheaf, with action on objects mapping $\Gamma$
to the set $A \in \Gamma$ of indices of $A$ in $\Gamma$ and action on morphisms
witnessed by $\treindex : \Gamma \sublist \Delta \to A \in \Gamma \to A \in
\Delta$.  The associated functor laws $\treindex\,\tid\,x \equiv x$ and
$\treindex\,\tau_2\,(\treindex\,\tau_1\,x) \equiv \treindex\,(\tau_1 \rcomp
\tau_2)\,x$ hold propositionally.

@@

Simple types shall be distinguished into positive types $P$ and
negative types $N$, depending on their root type former.  Function
($\To$) and product types ($\times$ and $\tyone$) are negative,
while base types ($\tybase$) and sum types ($+$ and $\tyzero$) are positive.
\[
\begin{array}{lrl@{\qquad}l}
  A,B,C & ::= & P \mid N & \mbox{simple types} \\
  P     & ::= & \tyzero \mid A + B \mid \tybase & \mbox{positive types} \\
  N     & ::= & \tyone \mid A \times B \mid A \To B
  % P     & ::= & a \mid A + B \mid \tyzero & \mbox{positive types} \\
  % N     & ::= & A \To B \mid A \times B \mid \tyone
    & \mbox{negative types} \\
\end{array}
\]
% Simple types shall be given by the grammar
% % \fbox{$S,T ::= P \mid S \To T \mid S \times T \mid \tyone \mid S + T
% %   \mid \tyzero$}.
% \fbox{$A,B,C ::= a \mid A \To B \mid A \times B \mid \tyone \mid A + B
%   \mid \tyzero$}.
% Herein, $a$ ranges over atomic types aka base types.
Intrinsically well-typed lambda-terms, in abstract syntax,
are just inhabitants $t$ of the indexed set \fbox{$A \from \Gamma$},
inductively defined by the following rules.
\begin{gather*}
  \nru{\tvar}{A \in \Gamma}{A \from \Gamma}
\quad
  \nru{\tabs}{B \from \cext \Gamma A}{A \To B \from \Gamma}
\quad
  \nru{\tapp}
    {A \To B \from \Gamma \quad A \from \Gamma}
    {B \from \Gamma}
\\[2ex]
  \nru{\tunit}{}{\tyone \from \Gamma}
\quad
  \nru{\tpair}
    {A_1 \from \Gamma \quad A_2 \from \Gamma}
    {A_1 \times A_2 \from \Gamma}
\quad
  \nru{\tprj_i}
    {A_1 \times A_2 \from \Gamma}
    {A_i \from \Gamma}
\\[2ex]
  \nru{\tinj_i}
    {A_i \from \Gamma}
    {A_1 + A_2 \from \Gamma}
\quad
  \nru{\tcase}
    {A_1 + A_2 \from \Gamma
      \quad B \from \cext \Gamma A_1
      \quad B \from \cext \Gamma A_2
    }{B \from \Gamma}
\\[2ex]
  \nru{\tabort}
    {\tyzero \from \Gamma}
    {B \from \Gamma}
\end{gather*}
The skilled eye of the reader will immediately recognize the proof
rules of intuitionistic propositional logic (IPL) under the
Curry-Howard isomorphism, where $A \from \Gamma$ is to be read as
``$A$ follows from $\Gamma$''.  Using shorthand
\fbox{$\vv n = \tvar\,(\tsuc^n\,\tzero)$} for the $n$th variable,
a term such as
\[
  \tabs\,(\tabs\,(\tpair\,
    \vv1\,
    (\tabs\,(\tapp\,\vv1\,\vv0))))
\]
could in concrete syntax be rendered as
\(
  \lambda x.\,\lambda y.\,(x,\,\lambda z.\,y\,z)
  .
\)
We leave the exact connection to a printable
syntax of the STLC to the imagination of the reader,
as we shall not be concerned with considering
concrete terms in this article.

@@

Normal forms%
\footnote{%
There is also a stronger
notion of normal form, requiring that two \emph{extensionally equal}
lambda-terms, \ie, those that denote the
same set-theoretical function, have the same normal form
\cite{mitchell:foundations,altenkirchDybjerHofmannScott:lics01,scherer:popl17}.
Such normal forms do not have a simple inductive definition, and we
shall not consider them in this article.
}
are logically characterized as those fulfilling the
\emph{subformula} property \cite{prawitz:natded,matthes:shortproofs}.
Normal forms \fbox{$n : \tNf A\, \Gamma$} are mutually defined with
\emph{neutral} normal forms \fbox{$u : \Ne A \Gamma$}.
In the following inductive definition, we reuse the rule names from the
term constructors.
\begin{gather*}
  \nru{\tvar}{A \in \Gamma}{\Ne A \Gamma}
\qquad
  \nru{\tprj_i}
    {\NeP {A_1 \times A_2} \Gamma}
    {\Ne {A_i} \Gamma}
\\[2ex]
  \nru{\tapp}
    {\NeP {A \To B} \Gamma \qquad \Nf A \Gamma}
    {\Ne B \Gamma}
\qquad
  \nru{\tne}{\Ne \tybase \Gamma}{\Nf \tybase \Gamma}
\\[2ex]
  \nru{\tunit}{}{\Nf \tyone \Gamma}
\qquad
  \nru{\tpair}
    {\Nf {A_1} \Gamma \qquad \Nf {A_2} \Gamma}
    {\NfP {A_1 \times A_2} \Gamma}
\\[2ex]
  \nru{\tabs}{\NfC B {\cext \Gamma A}}{\NfP {A \To B} \Gamma}
\qquad
  \nru{\tinj_i}
    {\Nf {A_i} \Gamma}
    {\NfP {A_1 + A_2} \Gamma}
\\[2ex]
  \nru{\tcase}
    {\NeP {A_1 + A_2} \Gamma
      \qquad \NfC P {\cext \Gamma A_1}
      \qquad \NfC P {\cext \Gamma A_2}
    }{\Nf P \Gamma}
\\[2ex]
  \nru{\tabort}
    {\Ne \tyzero \Gamma}
    {\Nf P \Gamma}
\end{gather*}
These rules only allow the elimination of \emph{neutrals}; this
restriction guarantees
the subformula property and prevents any kind of computational
($\beta$) redex.
The new rule $\tne$ embeds $\tNe$ into $\tNf$,
but only at base types $\tybase$ \cite[Section~3.3]{altenkirchHofmannStreicher:ctcs95}.
Further, case distinction via $\tcase$ and $\tabort$
is restricted to positive types $P$.
As a consequence, our normal forms are
\emph{$\eta$-long}, meaning that any normal inhabitant of a negative
type is a respective introduction ($\tabs$, $\tunit$, or $\tpair$).
This justifies the attribute \emph{negative} for these types:
the construction of their inhabitants proceeds mechanically, without
any choices.  In contrast, constructing an inhabitant of a
\emph{positive} type involves choice: whether case distinction is
required, and which introduction to pick in the end ($\tinj_1$ or $\tinj_2$).

@@

\begin{remark}
While the coproduct eliminations $\tcase$ and $\tabort$ are limited to
normal forms of positive types $P$, their extension $\tcase^B$ and
$\tabort^B$ to negative types
is admissible, for instance:
\[
\begin{array}{l@{~}lcl}
  \multicolumn 4 l {\tabort^B ~:~ \tNe\, 0 \todot \tNf\,B} \\
  \tabort^{\tyone}     & u & = & \tunit \\
  \tabort^P           & u & = & \tabort\;u \\
% \end{array}
% \qquad
% \begin{array}{l@{~}lcl}
%   \\
  \tabort^{A \times B} & u & = & \tpair\,(\tabort^A\,u)\,(\tabort^B\,u) \\
  \tabort^{A \To B}    & u & = & \tabs\,(\tabort^B\,(\tren\,\twk^A\,u)) \\
\end{array}
% \begin{array}{l@{~}lcl}
%   \multicolumn 4 l {\tabort^B ~:~ \tNe\, 0 \todot \tNf\,B} \\
%   \tabort^{\tyone}     & u & = & \tunit \\
%   \tabort^{A \times B} & u & = & \tpair\,(\tabort^A\,u)\,(\tabort^B\,u) \\
%   \tabort^{A \To B}    & u & = & \tabs\,(\tabort^B\,(\tren\,\twk^A\,u)) \\
%   \tabort^P           & u & = & \tabort\;u \\
% \end{array}
\]
$\tcase$ generalizes analogously, with a bit of care when weakening
the branches.
\end{remark}

@@

%\emph{Normalization}
Normalization is concerned with finding a normal form
$n : \Nf A \Gamma$ for each term $t : A \from \Gamma$.
The normal form should be \emph{sound},
\ie, $\tm n \cong t$
with respect to an equational theory $\cong$ on terms
(see \cref{sec:eq}).
Further, normalization should decide $\cong$, \ie,
terms $t,t'$ with $t \cong t'$ should have the same normal form $n$.
In this article, we implement only the normalization function
\fbox{$\tnorm : A \from \Gamma \to \Nf A \Gamma$}
with proving its soundness and completeness.
From a logical perspective, we will compute for each derivation of $A
\from \Gamma$ a normal derivation $\Nf A \Gamma$.

@@

% that does not guarantee any relation between its input term $t$ and
% output normal form $u$.
%
% \emph{Normalization by evaluation} (NbE) decomposes normalization as
% \fbox{\(
%   \norm[\Gamma] A (t : A \from \Gamma)
%   = \down[\Gamma] A \dent t {\idenv\Gamma}
% \)}
% which is
% \emph{evaluation}
% \nofbox{$\dent \_ \_ : (t : A \from \Gamma) \to \Den \Gamma \Delta \to
%   \Den A \Delta$}
% in the identity environment \nofbox{$\idenv\Gamma : \Den \Gamma \Gamma$}
% followed by
% \emph{reification}
% \nofbox{$\downG A : \Den A \Gamma \to \Nf A \Gamma$}.
\emph{Normalization by evaluation} (NbE)
%\fbox{\(
%  \norm[\Gamma] A (t : A \from \Gamma)
\[
  \tnorm\, (t : A \from \Gamma)
  = \down A \dent t {\idenv\Gamma}
\]
%\)}
decomposes normalization into
\emph{evaluation}
\nofbox{$\denty\_ : (t : A \from \Gamma) \to \den \Gamma \todot
  \den A$}
in the identity environment \nofbox{$\idenv\Gamma : \den \Gamma \Gamma$}
followed by
\emph{reification}
\nofbox{$\down A : \den A \todot \tNf A$}
(aka quoting).
The role of evaluation is to produce from a term the corresponding
semantic (\ie, meta-theoretic) function, which is finally reified to a normal form.
Since we are evaluating open terms $t$, we need to supply an environment
$\idenv\Gamma$ which will map the free indices of $t$ to corresponding
\emph{unknowns}.  To accommodate unknowns in the semantics, types $A$
are mapped to presheaves $\den A$ (rather than just sets), and in
particular each base type $\tybase$ is mapped to the presheaf
$\tNe\,\tybase$ with the intention that the neutrals take the role of
the unknowns.  The mapping $\up A : \tNe\,A \todot \den A$ from
neutrals to unknowns is called \emph{reflection} (aka unquoting), and defined mutually
with reification by induction on type $A$.

@@

Presheaves (co)products $\hat0$, $\hat1$, $\hatplus$, and $\hattimes$ are
constructed pointwise, \eg, $\hat0\,\Gamma = \bzero$, and given two
presheaves $\A$ and $\B$,
$(\A \hatplus \B)\,\Gamma = \A\,\Gamma \bplus \B\,\Gamma$.  For the
exponential of presheaves, however, we need the \emph{Kripke function
  space}
$(\A \hatto \B)\,\Gamma = \forall \Delta.\, \Gamma \sublist \Delta
\to \A\,\Delta \to \B\,\Delta$.

>>>>

% \hypersetup{draft}  %% debugging url pagebreaks

@@

%
% The "author" command and its associated commands are used to define the authors and their affiliations.
% Of note is the shared affiliation of the first two authors, and the "authornote" and "authornotemark" commands
% used to denote shared contribution to the research.
\author{Andreas Abel}
%\authornote{Both authors contributed equally to this research.}
%\email{andreas.abel@gu.se}
\orcid{0000-0003-0420-4492}
%\authornotemark[1]
\affiliation{%
  \institution{Department of Computer Science and Engineering, Chalmers and Gothenburg University}
  \streetaddress{Rännvägen 6b}
  \city{Göteborg}
  \state{Sweden}
  \postcode{41296}
}

\author{Christian Sattler}
\affiliation{%
  \institution{Nottingham University}
  \streetaddress{}
  \city{Nottingham}
  \country{UK}
}
%\email{christian.sattler@gmail.com}

@@

The idea behind \emph{normalization by evaluation} (NbE) is to utilize a
standard interpreter, usually evaluating closed terms, to compute the
normal form of an open term.
The normal form is obtained by a type-directed \emph{reification}
procedure after evaluating the open term to a semantic value, mapping
(\emph{reflecting})
the free variables to corresponding \emph{unknowns} in the semantics.
The literal use of a standard interpreter
can be achieved for the pure simply-typed lambda-calculus
\cite{bergerSchwichtenberg:lics91,filinski:semaccounttdpe}
by modelling uninterpreted base types as sets of neutral (aka atomic)
terms.   More precisely, base types are interpreted
as presheaves or sets of neutral term
families, in order to facilitate the generation of
fresh variables of base type during
reification of functions to lambdas.
Thanks to $\eta$-equality at function types, free variables of
\emph{function} type can be reflected into the semantics as functions
applying the variable to their reified argument,
forming a neutral term.  This mechanism provides us with unknowns of
base and function type which can be faithfully reified to normal forms.
% Since neutrals are closed under application to normal arguments, they
% can act as unknown functions, storing the reified semantic argument
% they are invoked with (thus, these unknowns are aka accumulators
% \cite{gregoireleroy:icfp02}).

For the extension to \emph{sum} types (logically, disjunctions, and categorically,
weak coproducts), this reflection trick does not work anymore.  A
semantic value of binary sum type is either a left or a right
injection, but the decision between left or right cannot be taken at
reflection time, since a variable of sum type does not provide us with
such information.  A literal standard interpreter for closed terms can
thus no longer be used for NbE;  instead, we can utilize a monadic
interpreter.  When the interpreter attempts a case distinction on an
unknown of sum type, it asks an oracle whether the unknown is a left
or a right injection.  The oracle returns one of these alternatives,
wrapping a new unknown in the respective injection.  The communication with the
oracle can be modeled in a \emph{monad} $\C$,
% with the service $\tNe\,(A + B) \to \C\,(\den A + \den B)$
which records the questions asked and the continuation of the
interpreter for each of the possible answers.
A monadic semantic value is thus a \emph{case tree} where the leaves
are occupied by non-monadic values \cite{altenkirchUustalu:flops04}.
In this article we only consider weak sum types, producing non-unique
normal forms.  In particular, case trees are not normalized, allowing,
\eg, redundant case splits (\ie, asking the same question twice).
Further, the order of case splits is not normalized: changing the
order in which the questions are asked (commuting case splits),
alters the normal form.
The model would need refinement for extensional sums (strong
coproducts) with unique normal forms
\cite{altenkirchDybjerHofmannScott:lics01,altenkirchUustalu:flops04,balatDiCosmoFiore:popl04,barral:PhD,scherer:popl17}.

@@

In this work, we formulate NbE for CBPV
(Section~\ref{sec:cbpv}).
% , with the aim to investigate
% later whether CBN and CBV NbE can be recovered from CBPV NbE via the
% standard translations of the CBN and CBV calculi into CBPV.
%
In contrast to the normal forms of CBN NbE, which is the algorithmic
counterpart of the completeness proof for intuitionistic propositional
logic (IPL) using Beth models, CBPV NbE gives us more restrained
normal forms, where the production of a value via injections cannot be
interrupted by more questions to the oracle.  In the research field of
focalization \cite{andreoli:focusing,liangMiller:csl07} we speak of
\emph{chaining non-invertible introductions}.
Invertible introductions are already chained in NbE thanks to
extensionality ($\eta$) for function, and more generally, negative
types.  Non-invertible eliminations are also happening in a chain when
building neutrals.  What is missing from the picture is the chaining
of invertible eliminations, i.e., case distinctions and, more
generally, pattern matching.  The picture is completed by extending
NbE to \emph{polarized lambda calculus}
\cite{zeilberger:PhD,brockNannestadSchuermann:lpar10,espiritoSanto:entcs17}
in Section~\ref{sec:fipl}.

@@

In category theory, a presheaf $\A$ over category $\bC$ is a contravariant
functor from $\bC$ into the category $\Set$ of sets and functions, \ie, a
(covariant) functor from the category $\bC\op$ with flipped morphisms
to $\Set$.
In this paper, we are only interested in (covariant) functors from $\Cxt$ to
$\Set$, thus,
by a \emph{presheaf} we will always mean a presheaf over the category $\Cxt\op$.
% i.e., a functor from $\Cxt$ to $\Set$.
% \footnote{With respect to the terms
% defined later, $\Cxt\op$ is to be seen as a category of contexts and those
% substitutions given by weakening.}
For example, given any type $A$, we can
consider $A{\in}\_$ as such a presheaf, with the action on objects mapping $\Gamma$
to the set $A \in \Gamma$ of indices of $A$ in $\Gamma$, and the action on morphisms
% witnessed by
being
$\treindex : \Gamma \sublist \Delta \to A \in \Gamma \to A \in
\Delta$.  The associated functor laws $\treindex\,\tid\,x \equiv x$ and
$\treindex\,\tau_2\,(\treindex\,\tau_1\,x) \equiv \treindex\,(\tau_1 \rcomp
\tau_2)\,x$ hold propositionally.

@@

Simple types shall be distinguished into positive types $P$ and
negative types $N$, depending on their root type former (for now).  Function
($\To$) and product types ($\times$ and $\tyone$) are negative,
while base types ($\tybase$) and sum types ($+$ and $\tyzero$) are positive.
\[
\begin{array}{lrl@{\qquad}l}
  A,B,C & ::= & P \mid N & \mbox{simple types} \\
  P     & ::= & \tyzero \mid A + B \mid \tybase & \mbox{positive types} \\
  N     & ::= & \tyone \mid A \times B \mid A \To B
  % P     & ::= & a \mid A + B \mid \tyzero & \mbox{positive types} \\
  % N     & ::= & A \To B \mid A \times B \mid \tyone
    & \mbox{negative types} \\
\end{array}
\]
% Simple types shall be given by the grammar
% % \fbox{$S,T ::= P \mid S \To T \mid S \times T \mid \tyone \mid S + T
% %   \mid \tyzero$}.
% \fbox{$A,B,C ::= a \mid A \To B \mid A \times B \mid \tyone \mid A + B
%   \mid \tyzero$}.
% Herein, $a$ ranges over atomic types aka base types.
Intrinsically well-typed lambda-terms, in abstract syntax,
are just inhabitants $t$ of the indexed set \fbox{$A \from \Gamma$},
inductively defined by the following rules.
\begin{gather*}
  \nru{\tvar}{A \in \Gamma}{A \from \Gamma}
\quad
  \nru{\tabs}{B \from \cext \Gamma A}{A \To B \from \Gamma}
\quad
  \nru{\tapp}
    {A \To B \from \Gamma \quad A \from \Gamma}
    {B \from \Gamma}
\\[2ex]
  \nru{\tunit}{}{\tyone \from \Gamma}
\quad
  \nru{\tpair}
    {A_1 \from \Gamma \quad A_2 \from \Gamma}
    {A_1 \times A_2 \from \Gamma}
\quad
  \nru{\tprj_i}
    {A_1 \times A_2 \from \Gamma}
    {A_i \from \Gamma}
\\[2ex]
  \nru{\tinj_i}
    {A_i \from \Gamma}
    {A_1 + A_2 \from \Gamma}
\quad
  \nru{\tcase}
    {A_1 + A_2 \from \Gamma
      \quad B \from \cext \Gamma A_1
      \quad B \from \cext \Gamma A_2
    }{B \from \Gamma}
\\[2ex]
  \nru{\tabort}
    {\tyzero \from \Gamma}
    {B \from \Gamma}
\end{gather*}
The skilled eye of the reader will immediately recognize the proof
rules of intuitionistic propositional logic (IPL) under the
Curry-Howard isomorphism, where $A \from \Gamma$ is to be read as
``$A$ follows from $\Gamma$''.  Using shorthand
\fbox{$\vv n = \tvar\,(\tsuc^n\,\tzero)$} for the $n$th variable,
a term such as
\[
  \tabs\,(\tabs\,(\tpair\,
    \vv1\,
    (\tabs\,(\tapp\,\vv1\,\vv0))))
\]
could in concrete syntax be rendered as
\(
  \lambda x.\,\lambda y.\,(x,\,\lambda z.\,y\,z)
  .
\)
We leave the exact connection to a printable
syntax of the STLC to the imagination of the reader,
as we shall not be concerned with considering
concrete terms in this article.

@@

Normal forms%
\footnote{%
There is also a stronger
notion of normal form, requiring that two \emph{extensionally equal}
lambda-terms, \ie, those that denote the
same set-theoretical function, have the same normal form
\cite{mitchell:foundations,altenkirchDybjerHofmannScott:lics01,scherer:popl17}.
Such normal forms do not have a simple inductive definition, and we
shall not consider them in this article.
}
are logically characterized as those fulfilling the
\emph{subformula} property \cite{prawitz:natded,matthes:shortproofs}.
Normal forms \fbox{$n : \tNf A\, \Gamma$} are mutually defined with
\emph{neutral} normal forms \fbox{$u : \Ne A \Gamma$}.
In the following inductive definition, we reuse the rule names from the
term constructors.
\begin{gather*}
  \nru{\tvar}{A \in \Gamma}{\Ne A \Gamma}
\qquad
  \nru{\tprj_i}
    {\NeP {A_1 \times A_2} \Gamma}
    {\Ne {A_i} \Gamma}
\\[2ex]
  \nru{\tapp}
    {\NeP {A \To B} \Gamma \qquad \Nf A \Gamma}
    {\Ne B \Gamma}
\qquad
  \nru{\tne}{\Ne \tybase \Gamma}{\Nf \tybase \Gamma}
\\[2ex]
  \nru{\tunit}{}{\Nf \tyone \Gamma}
\qquad
  \nru{\tpair}
    {\Nf {A_1} \Gamma \qquad \Nf {A_2} \Gamma}
    {\NfP {A_1 \times A_2} \Gamma}
\\[2ex]
  \nru{\tabs}{\NfC B {\cext \Gamma A}}{\NfP {A \To B} \Gamma}
\qquad
  \nru{\tinj_i}
    {\Nf {A_i} \Gamma}
    {\NfP {A_1 + A_2} \Gamma}
\\[2ex]
  \nru{\tcase}
    {\NeP {A_1 + A_2} \Gamma
      \qquad \NfC P {\cext \Gamma A_1}
      \qquad \NfC P {\cext \Gamma A_2}
    }{\Nf P \Gamma}
\\[1ex]
  \nru{\tabort}
    {\Ne \tyzero \Gamma}
    {\Nf P \Gamma}
\end{gather*}
These rules only allow the elimination of \emph{neutrals}; this
restriction guarantees
the subformula property and prevents any kind of computational
($\beta$) redex.
The new rule $\tne$ embeds $\tNe$ into $\tNf$,
but only at base types $\tybase$ \cite[Section~3.3]{altenkirchHofmannStreicher:ctcs95}.
Further, case distinction via $\tcase$ and $\tabort$
is restricted to positive types $P$.
As a consequence, our normal forms are
\emph{$\eta$-long}, meaning that any normal inhabitant of a negative
type is a respective introduction ($\tabs$, $\tunit$, or $\tpair$).
This justifies the attribute \emph{negative} for these types:
the construction of their inhabitants proceeds mechanically, without
any choices.  In contrast, constructing an inhabitant of a
\emph{positive} type involves choice: whether case distinction is
required, and which introduction to pick in the end ($\tinj_1$ or $\tinj_2$).

@@

We use the \fbox{$\forall$} quantifier as implicit dependent function
type former in our meta-language,
similar to the polymorphic quantifier in functional
programming languages such as Haskell.  We may instantiate a
polymorphic function such as $f : \A \todot \B$ silently, \eg, when $a
: \A\,\Gamma$ then $f\,a : \B\,\Gamma$.  For clarity, we may sometimes
provide the instantiation explicitely via subscript, \eg, $f_\Gamma :
\A\,\Gamma \to \B\,\Gamma$ and $f_\Gamma\,a : B\,\Gamma$.  Similarly,
to introduce a polymorphic function we may provide the implicit
abstraction as subscript, such as in
$\lambda\,{}_\Gamma\, a.\ b : \A \todot \B$ or
$\lambda\,{}_\Gamma\, (a : \A\,\Gamma).\ b : \A \todot \B$, or
omit it, \eg,
$\lambda\,(a : \A\,\Gamma).\ b : \A \todot \B$.

\begin{remark}
While the coproduct eliminations $\tcase$ and $\tabort$ are limited to
normal forms of positive types $P$, their extension $\tcase^B$ and
$\tabort^B$ to negative types
is admissible, for instance:
\[
\begin{array}{l@{~}lcl}
  \multicolumn 4 l {\tabort^B ~:~ \tNe\, 0 \todot \tNf\,B} \\
  \tabort^{\tyone}     & u & = & \tunit \\
  \tabort^P           & u & = & \tabort\;u \\
% \end{array}
% \qquad
% \begin{array}{l@{~}lcl}
%   \\
  \tabort^{A_1 \times A_2} & u & = & \tpair\,(\tabort^{A_1}\,u)\,(\tabort^{A_2}\,u) \\
  \tabort^{A \To B}    & u & = & \tabs\,(\tabort^B\,(\tren\,\twk^A\,u)) \\
\end{array}
% \begin{array}{l@{~}lcl}
%   \multicolumn 4 l {\tabort^B ~:~ \tNe\, 0 \todot \tNf\,B} \\
%   \tabort^{\tyone}     & u & = & \tunit \\
%   \tabort^{A \times B} & u & = & \tpair\,(\tabort^A\,u)\,(\tabort^B\,u) \\
%   \tabort^{A \To B}    & u & = & \tabs\,(\tabort^B\,(\tren\,\twk^A\,u)) \\
%   \tabort^P           & u & = & \tabort\;u \\
% \end{array}
\]
$\tcase$ generalizes analogously, with a bit of care when weakening
the branches.
\end{remark}

@@

%\emph{Normalization}
Normalization is concerned with finding a normal form
$n : \Nf A \Gamma$ for each term $t : A \from \Gamma$.
The normal form should be \emph{sound},
\ie, $\tm n \cong t$
with respect to an equational theory $\cong$ on terms%
. %(see \cref{sec:eq}).
Further, normalization should decide $\cong$, \ie,
terms $t,t'$ with $t \cong t'$ should have the same normal form $n$.
In this article, we implement only the normalization function
\fbox{$\tnorm : A \from \Gamma \to \Nf A \Gamma$}
with proving its soundness and completeness.
From a logical perspective, we will compute for each derivation of $A
\from \Gamma$ a normal derivation $\Nf A \Gamma$.

@@

% that does not guarantee any relation between its input term $t$ and
% output normal form $u$.
%
% \emph{Normalization by evaluation} (NbE) decomposes normalization as
% \fbox{\(
%   \norm[\Gamma] A (t : A \from \Gamma)
%   = \down[\Gamma] A \dent t {\idenv\Gamma}
% \)}
% which is
% \emph{evaluation}
% \nofbox{$\dent \_ \_ : (t : A \from \Gamma) \to \Den \Gamma \Delta \to
%   \Den A \Delta$}
% in the identity environment \nofbox{$\idenv\Gamma : \Den \Gamma \Gamma$}
% followed by
% \emph{reification}
% \nofbox{$\downG A : \Den A \Gamma \to \Nf A \Gamma$}.
\emph{Normalization by evaluation} (NbE)
%\fbox{\(
%  \norm[\Gamma] A (t : A \from \Gamma)
\[
  \tnorm\, (t : A \from \Gamma)
  = \down A \dent t {\idenv\Gamma}
\]
%\)}
decomposes normalization into
\emph{evaluation}
\[%\nofbox{$
\denty\_ : (t : A \from \Gamma) \to \den \Gamma \todot
  \den A
\]%$}
in the identity environment
\[%\nofbox{$
\idenv\Gamma : \den \Gamma \Gamma
\]%$}
followed by
\emph{reification}
\[%\nofbox{$
\down A : \den A \todot \tNf A
\]%$}
(aka quoting).
The role of evaluation is to produce from a term the corresponding
semantic (\ie, meta-theoretic) function, which is finally reified to a normal form.
Since we are evaluating open terms $t$, we need to supply an environment
$\idenv\Gamma$ which will map the free indices of $t$ to corresponding
\emph{unknowns}.  To accommodate unknowns in the semantics, types $A$
are mapped to presheaves $\den A$ (rather than just sets), and in
particular each base type $\tybase$ is mapped to the presheaf
$\tNe\,\tybase$ with the intention that the neutrals take the role of
the unknowns.  The mapping $\up A : \tNe\,A \todot \den A$ from
neutrals to unknowns is called \emph{reflection} (aka unquoting), and defined mutually
with reification by induction on type $A$.

@@

Presheaf (co)products $\hat0$, $\hat1$, $\hatplus$, and $\hattimes$ are
constructed pointwise, \eg, $\hat0\,\Gamma = \bzero$, and given two
presheaves $\A$ and $\B$,
$(\A \hatplus \B)\,\Gamma = \A\,\Gamma \bplus \B\,\Gamma$.  For the
exponential of presheaves, however, we need the \emph{Kripke function
  space}
$(\A \hatto \B)\,\Gamma = \forall \Delta.\, \Gamma \sublist \Delta
\to \A\,\Delta \to \B\,\Delta$.
